Examining equity in accessibility to bike share - a balanced floating catchment area approach

### 2021-06-18 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Major revisions of manuscript

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining horizontal and vertical equity to a public bicycle share
program using a balanced floating catchment area accessibility approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Public bicycle share programs (PBSPs) can play a role in advancing
  transportation equity if they make cycling more accessible to
  disadvantaged populations. In Ontario, the City of Hamilton expanded
  their PBSP in 2018 by adding twelve docking stations with the explicit
  objective of increasing spatial equity in access. In this case study, we
  investigate differentials in accessibility to stations using a balanced
  floating catchment area accessibility approach and compare accessibility
  with and without the equity stations. We analyze micro zones to better
  reflect walking to a station and conduct a sensitivity analysis at
  several walking time thresholds. We then reaggregate the estimated
  accessibility for further analysis using census data. Our findings
  indicate that equity stations increased the serviced population at every
  threshold examined. Although accessibility increased for the whole
  population, this increase was relatively modest especially for
  population in the bottom 20\% of median total household income.
  \end{abstract}

@@

@@

The potential of public bicycle share programs (PBSPs) to increase
cycling levels is but one of many reasons for implementing such programs
in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
and convenient form of public transportation, shared bicycles can
encourage individuals to take up cycling for short local trips or first
and last mile trips to other public transportation instead of using
personal vehicles. These programs can also play a role in addressing
transportation needs and advancing transportation equity if they make
cycling more accessible to individuals and groups with lower
socioeconomic status. As these programs become increasingly common,
their introduction has been accompanied by a flurry of research that
investigates them from the perspective of spatial and transportation
equity (e.g., Hosford and Winters, 2018; Hull Grasso et al., 2020;
Mooney et al., 2019; Qian and Jaller, 2021, 2020; Smith et al., 2015).
Indeed, although these programs are available to the general public and
ought to be accessible to any individual who wishes to use them,
research on PBSPs located in North American and European cities
indicates that inequity in spatial accessibility needs redressing.

In this paper, we investigate the case of the PBSP in the city of
Hamilton, in Ontario, Canada. Hamilton Bike Share was launched in 2015
and currently has over 900 operational bicycles and 130 docking
stations. An interesting feature of this program is that it was expanded
by an equity initiative that introduced twelve stations in 2018 with the
explicit objective of increasing spatial accessibility. Hamilton Bike
Share was already studied by Hosford and Winters (2018) as part of a
selection of PBSPs in Canada. The authors found that most of the cities
in their research could benefit of greater efforts to expand service to
areas with lower socioeconomic groups - but Hamilton fared somewhat
better than other cities in this respect. What distinguishes our
research from existing literature is the use of a balanced floating
catchment area (BFCA) accessibility approach that accounts for the
supply of stations and the potential demand from the population
serviced. This requires a more disaggregated approach than the use of
census geographies (q.v., Hosford and Winters, 2018), since walking
trips to a bicycle share station likely happen at a level lower than
even the smallest census geography. For this reason, we implement our
analysis parting from micro population zones to better reflect the
friction of walking to a docking station, which is an important
component of a bike share trip (Chen et al., 2019). Further, we conduct
a sensitivity analysis at several walking time thresholds. In terms of
spatial equity, we compare accessibility with and without the equity
stations to assess the effect of the initiative, before reaggregating
the data for further analysis using median total household income
information from the census.

@@

\hypertarget{literature-review}{%
\section{2. Literature Review}\label{literature-review}}

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
However, one study found that bike share users in Washington, DC were
more likely to be female (Buck et al., 2013), which suggests that the
gender gap among cyclists who use bike share is less disparate than the
gap for personal bicycle use (Fishman, 2016). There is some evidence
that bike share users are less likely to own a car (Buck et al., 2013;
Reilly, Noyes, et al., 2020). However, the relationship between income
or education and bike share use is less clear-cut. Stations in
disadvantaged communities in Chicago have been found to generate most of
the average annual trips (Qian and Jaller, 2020) and individuals from
minority or lower socioeconomic status neighborhoods in
Minneapolis-St.~Paul used the city's PBSP more (J. Wang and Lindsey,
2019a). Similar findings were reported in London (Ogilvie and Goodman,
2012). Being university educated was a significant correlate of bike
share use in Montreal, Canada (Fuller et al., 2011). Perhaps not
coincidentally, financial savings have been found to motivate those on a
low income to use bike share (Fishman, 2016).

Many studies have found that geographic proximity to a bicycle share
station is an important determinant of membership and use (Fishman,
2016; Fuller et al., 2011). This makes sense given that individuals are
more likely to use services or programs that they can easily reach.
Several studies have recently explored the equity of PBSPs in North
America by primarily examining who has access to bike share (e.g.,
differences by demographics or socioeconomic status) and where stations
are located. It is important to note that equity can be achieved in two
different ways: balanced or equal spatial distribution across a region
for all similar groups (e.g., horizontal equity) or greater and targeted
access to accommodate vulnerable or disadvantaged populations (e.g.,
vertical equity) (see Chen et al., 2019). Both are of interest to
researchers and transportation planners since they are often linked in
that advantage, or conversely disadvantage, has spatial patterns. Using
a negative binomial regression model, Qian and Jaller (2020) estimated
ridership in Chicago's PBSP and found some disparities. A minority of
bike share stations were found to be located in disadvantaged
communities, while annual members from disadvantaged communities have a
lower share of trips compared to other areas in the city. This suggests
that they may be using PBSPs for utilitarian trips (e.g., commuting),
which points to the importance of ensuring equitable access (Qian and
Jaller, 2020). Similar results were found in Philadelphia. Despite
efforts to increase equity within the city's PBSP, census block groups
with lower median income generated fewer trips which suggested that
efforts have not been as successful as intended (Caspi and Noland,
2019). Trips from stations in such areas are utilitarian (e.g.,
commuting to work), which points to the importance of ensuring equitable
access (Caspi and Noland, 2019). In the case of Seattle, all
neighborhoods had some level of access to dockless bicycles but those
with higher incomes and more residents of higher education had more
bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

On the whole, existing studies highlight the need for PBSPs to be more
highly accessible for diverse populations in order to increase use
beyond the ``typical'' users. This has been the focus of recent research
(see, among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. A few North American cities have launched
specific programs to address barriers to bike share or to expand service
to more deprived areas: Philadelphia, Pennsylvania (Caspi and Noland,
2019), and Hamilton, Ontario (Hosford and Winters, 2018). However,
exploring transportation equity by investigating where bicycle share
stations are located, often using neighborhoods or census tracts as the
geographical unit of analysis, can ignore or miss the benefits that may
be derived from adjacent zones. Meaning that, stations may be lacking in
certain neighborhoods but there may be stations accessible within a
reasonable walking time. This is where geographical accessibility
becomes an important consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved over the past century and has increasingly become
linked with recent planning interests in prioritizing modes that are
suitable for local trips like walking and cycling (Levine, 2020). As a
measure of the ease of reaching potential destinations spread spatially
in a given area, accessibility is relevant to PBSPs because it can
identify current inequality in the provision of infrastructure, as well
as guide interventions that increase access for groups that are
under-serviced or address gaps in transportation options. It also
addresses some of the challenges of other performance measures such as
level of service within a transportation network by measuring
person-based indicators and exploring differences in use between
population groups (Páez et al., 2012).

Beyond the utility derived from using shared bikes to destinations of
value, an important aspect of accessibility associated with PBSPs is the
distance an individual must travel to reach a bicycle share station
(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
distance needed to reach a bicycle share station decreases the potential
of accessing the program and ultimately their use of the program, the
location and size (e.g., number of bicycle racks) of stations matters.
Indeed, distance to bicycle share stations is associated with use
(Fuller et al., 2011; J. Wang and Lindsey, 2019b) and can be a barrier
to using PBSPs (Fishman et al., 2014). Kabra et al.~(2020) found that
the majority of bike share usage in Paris comes from areas within 300 m
of stations, which amounts to 2-4 minutes walking by an able-bodied
adult. Furthermore, similar to other public amenities affected by
crowding, like health care services (e.g., Pereira, Carlos Kauê Vieira
Braga, et al., 2021), the utility of stations is limited by the
\emph{maximum} number of bicycles that they can hold. The program may
not necessarily be improved if stations are easy to reach but offer only
a small number of bicycles. Likewise, more people may not opt to use the
program if the supply of bicycles available at nearby stations is
insufficient to meet demand. The location and size of stations is
important to increase the utility of this public transportation option
for more people, thus achieving vertical equity. Accessibility analyses
for PBSPs constitute a positive and evaluation-based approach that also
has the potential to inform equity efforts. For instance, Wang and
Lindsey (2019b) investigated whether new or relocated bicycle share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

Several approaches have been commonly used for measuring place-based
accessibility, including cumulative opportunities, gravity, and
utility-based measures (Handy and Niemeier, 1997). Geurs and Van Wee
(2004) and Paez et al.~(2012) provide recent overviews of various
formulations and applications of accessibility in transportation
research. The common gravity-based approach for example involves
weighting destination opportunities, such as the quantity of bicycle
share stations, by the time required to reach them from an origin using
an impedance function (Handy and Niemeier, 1997; Kwan, 1998). However,
while such measures are suitable for capturing the potential for
reaching destinations from a given location, they do not take demand or
congestion effects into account.

In contrast, floating catchment area (FCA) methods incorporate
information on capacity and demand in calculating accessibility. FCA
measures have been widely employed in healthcare accessibility research
and are typically calculated across two steps. In the first, a ratio of
supply to demand at service locations is calculated, such as the number
of beds at a hospital divided by the number of people within the
catchment area of the hospital, weighted by the distance involved in
reaching the facility. Next, these service level ratios are allocated
back to the population centers and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. While there have been many
methodological innovations in FCA methods (for example, Delamater, 2013;
Luo and Qi, 2009; Luo and Wang, 2003; Radke and Mu, 2000; Wan et al.,
2012), a recent improvement to this approach was achieved through a
simple and intuitive balancing of the impedance that addressed the
effects of demand and service inflation found in earlier FCA approaches
(see Paez et al., 2019).

When measuring accessibility, researchers have taken different
approaches when it comes to aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequity in PBSP usage. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

Using the balanced floating catchment area method (Paez et al., 2019), a
novel approach that has not been used yet in cycling research, with
disaggregated population data, we examine accessibility to a public
bicycle share program in Hamilton, Ontario. In this paper, we (1)
conduct a sensitivity analysis by measuring accessibility and level of
service to bicycle share stations at different walking time thresholds
to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
(2) explore the contribution of specific stations that were added to
Hamilton's PBSP to reducing both horizontal and vertical inequities; and
(3) examine whether disparities in accessibility exist according to
median total household income of dissemination areas within the core
service area.

@@

The focus of this study is the city of Hamilton, located in Ontario,
Canada. The city launched a public bicycle share program in March 2015
with 115 stations and 750 bicycles (Hamilton, 2015a). Before June 2020,
the program was known as Social Bicycles or SoBi Hamilton, but is now
called Hamilton Bike Share. Stations are spaced between 300 and 600 m
apart (Scott et al., 2021). The core service area spans 40 sq.km of the
city although it was planned to be 20 sq.km (Hamilton, 2015b), and
roughly 138,000 people are within 30 minutes walking of a bike share
station {[}see Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This
represents roughly one fifth of the total population of the Hamilton
Census Metropolitan Area according to the 2016 Canadian Census. The City
of Hamilton undertook a large public engagement campaign to validate the
locations of bicycle share stations that had been selected and to
crowdsource potential locations for additional locations (Hamilton,
2014). Most of the suggested locations were in the east end of the core
service area that lack stations or in neighborhoods not serviced by the
PBSP. The program was enthusiastically welcomed in the city in 2015 -
within three weeks of launching, 10,000 trips had been made (Hamilton,
2015a), however inadequate coverage given the size of Hamilton Bike
Share's service area was identified as a problem early on, and
transportation planners noted that the small size (i.e., supply of
bicycle racks) and low quantity of stations would lead to challenges in
balancing the system (Hamilton, 2015b). Based on public feedback,
thirteen stations were added to the original system.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining equity in accessibility to bike share - a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining equity in accessibility to bike share - a balanced floating
catchment area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  Public bicycle share programs (PBSPs) can play a role in advancing
  transportation equity if they make bicycling more accessible to
  disadvantaged populations. In Ontario, the City of Hamilton expanded
  their PBSP in 2018 by adding twelve ``equity'' stations with the
  explicit objective of increasing access for under-serviced
  neighbourhoods. In this case study, we investigate differentials in
  accessibility to stations using a balanced floating catchment area
  approach and compare accessibility with and without the equity stations.
  We analyze micro zones to better reflect walking to a station and
  conduct a sensitivity analysis at several walking time thresholds. We
  then reaggregate the estimated accessibility for further analysis using
  census data. Our findings indicate that equity stations increased
  accessibility for the serviced population at every threshold examined,
  but the increase was relatively modest especially for population in the
  bottom 20\% of median total household income.
  \end{abstract}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{rm}\NormalTok{(}\DataTypeTok{list =} \KeywordTok{ls}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

@@

The potential of public bicycle share programs (PBSPs) to increase
bicycling levels is but one of many reasons for implementing such
programs in urban areas (see Hosford et al., 2018, 2019). As a healthy,
inexpensive, and convenient form of public transportation, shared
bicycles can encourage individuals to take up bicycling for short local
trips or first and last mile trips to other public transportation
instead of using personal vehicles. These programs can also play a role
in advancing transportation equity if they make bicycling more
accessible to disadvantaged populations. Although PBSPs are available to
the general public in over 800 cities worldwide (Fishman, 2016), and
ought to be accessible to any individual who wishes to use them,
research on PBSPs indicates that inequities persist with respect to who
can use and access them.

Many PBSPs in North America now offer specific programs to address
equity that primarily focus on removing cost barriers and increasing
access for groups that are under-represented among existing users
(McNeil et al., 2019). The locations of docking stations is a common
consideration for increasing equity (Howland et al., 2017). Hamilton
Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian
PBSP included in a North American scan of bike share equity programs
(see McNeil et al., 2019). HBS was launched in 2015 and currently has
over 900 operational bicycles and 130 docking stations. An equity
program, Everyone Rides Initiative, was implemented in 2018 which
expanded the program by introducing twelve ``equity'' stations which
expanded HBS to more disadvantaged neighbourhoods in the core service
area.

This paper examines accessibility to Hamilton Bike Share using a
balanced floating catchment area (BFCA) approach. We also conduct a
comparative analysis to the conventional two-step floating catchment
area approach to highlight the benefits of this method which, to our
knowledge, has not yet been used in the cycling literature. The paper
also assesses the contribution of HBS's equity stations to reducing
inequities in accessibility for different groups according to median
total household income, and provides policy recommendations to further
improve equity.

@@

\hypertarget{literature-review}{%
\section{2.1. Literature Review}\label{literature-review}}

\hypertarget{public-bicycle-share-programs}{%
\subsection{1.1. Public Bicycle Share
Programs}\label{public-bicycle-share-programs}}

Public bicycle share programs have been implemented in over 800 cities
worldwide and a great deal has been learned about their typical users
(Fishman, 2016). In many cities, males use bike share more than females
(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
However, one study found that bike share users in Washington, DC were
more likely to be female (Buck et al., 2013), which suggests that the
gender gap among bicyclists who use PBSPs is less disparate than the gap
for personal bicycle use (Fishman, 2016). There is some evidence that
bike share users are less likely to own a car (Buck et al., 2013;
Reilly, Noyes, et al., 2020). However, the relationship between income
or education and bike share use is less clear-cut. Stations in
disadvantaged communities in Chicago have been found to generate most of
the average annual trips (Qian and Jaller, 2020) and individuals from
minority or lower socioeconomic status neighborhoods in
Minneapolis-St.~Paul also used the city's PBSP more (J. Wang and
Lindsey, 2019a). Similar findings were reported in London (Ogilvie and
Goodman, 2012). Being university educated was a significant correlate of
bike share use in Montreal, Canada (Fuller et al., 2011). Not
coincidentally, financial savings have been found to motivate those on a
low income to use bike share (Fishman, 2016).

\hypertarget{equity-of-pbsps}{%
\subsection{1.2. Equity of PBSPs}\label{equity-of-pbsps}}

The introduction of PBSPs has been accompanied by a flurry of research
focusing on benefits from them. Many studies have examined differences
in demographics or socioeconomic status between those who use or have
access to PBSPs and those who don't, while other research has explored
spatial inequities in where stations are located (e.g., Hosford and
Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
(2020) estimated ridership in the city's PBSP and found that a minority
of bike share stations were located in disadvantaged communities, while
annual members from such areas had a lower share of trips compared to
other areas in the city. Similar results were found in Philadelphia.
Despite efforts to increase equity within the city's PBSP, census block
groups with lower median income generated fewer trips (Caspi and Noland,
2019). Trips from stations in such areas were utilitarian (e.g.,
commuting to work), which points to the importance of ensuring equitable
access (Caspi and Noland, 2019). In the case of Seattle, all
neighborhoods had some level of access to dockless bicycles but those
with higher incomes and more residents of higher education had more
bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
neighborhoods in New York City with higher affluence had the greatest
proportion of Citi Bike stations.

Chen et al.~(2019) distinguish two types of equity related to bike
share, horizontal and vertical, based on the work of Delbosc and Currie
(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
distribution of accessibility and costs for all similar groups, while
vertical equity would involve greater and targeted access for
under-represented or disadvantaged populations (see Chen et al., 2019).
Both are of interest to researchers and transportation planners since
they are often linked in that advantage, or conversely disadvantage, has
spatial patterns. Bike share equity programs, as identified by McNeil et
al.~(2019), can be considered efforts to improve vertical equity since
they favour groups that have benefited less from PBSPs.

On the whole, existing studies highlight the need for PBSPs to be more
accessible for diverse populations in order to increase use beyond the
``typical'' users. This has been the focus of recent research (see,
among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. However, exploring transportation equity by
investigating where bicycle share stations are located, often using
neighborhoods or census tracts as the geographical unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighborhoods but there
may be stations accessible within a reasonable walking time. This is
where geographical accessibility becomes an important consideration.

\hypertarget{accessibility-approaches}{%
\subsection{2.3. Accessibility
Approaches}\label{accessibility-approaches}}

Accessibility has been applied in both a positive and normative way to
inform transportation planning (Páez et al., 2012), but its utility to
this field has evolved over the past century and has increasingly become
linked with recent planning interests in prioritizing modes that are
suitable for local trips like walking and cycling (Levine, 2020). Beyond
the utility derived from using shared bicycles to access destinations of
value, docking stations themselves are amenities because they offer a
transportation service. Therefore, the ease of reaching these stations,
which are spread spatially in a given area, can affect use of the PBSP.

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location of stations matters. Kabra et al.~(2020) found that the
majority of bike share usage in Paris comes from areas within 300 m of
stations, which amounts to 2-4 minutes walking by an adult who does not
have a disability. Similar to other public amenities affected by
crowding, the utility of stations is also limited by the \emph{maximum}
number of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bicycle share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

Several approaches have been commonly used for measuring place-based
accessibility, including cumulative opportunities, gravity, and
utility-based measures (Handy and Niemeier, 1997). The gravity-based
approach involves weighting destination opportunities, such as the
quantity of bicycle share stations, by the time required to reach them
from an origin using an impedance function (Handy and Niemeier, 1997;
Kwan, 1998). While such measures are suitable for capturing the
potential for reaching destinations from a given location, they do not
take demand or congestion effects into account which is an important
consideration when calculating accessibility for amenities such as bike
share stations.

In contrast, floating catchment area (FCA) methods have been widely
employed in health care accessibility research. The benefit of this
approach is that the method incorporates information on capacity,
demand, and congestion in calculating accessibility. This approach is
more appropriate and informative than calculating provider-to-population
ratios (PPR) that simply divide the level of supply of a service (e.g.,
the number of bicycle racks at a station) by the population who have
access to the service (Paez et al., 2019). In particular, the Two-Step
Floating Catchment Area (2SFCA) method (Luo and Wang, 2003; Radke and
Mu, 2000) produces flexible catchment areas instead of using rigid
boundaries like PPR. In the first step, a ratio of supply to demand at
service locations is calculated, such as the number of beds at a
hospital divided by the number of people within the catchment area of
the hospital, weighted by the distance involved in reaching the
facility. Next, these service level ratios are allocated back to the
population centers and summarized as a measure of congested
accessibility. Thus, this model does a good job of considering potential
crowding or competition for services. However, overlapping catchment
areas from the conventional FCA approach lead to inflation of population
totals and deflation of service levels across a study area which
generates inaccurate or misleading accessibility estimates (Paez et al.,
2019). While there have been many methodological innovations in FCA
methods (e.g., Delamater, 2013; Luo and Qi, 2009; Luo and Wang, 2003;
Radke and Mu, 2000; Wan et al., 2012), a recent improvement to this
approach was achieved through a simple and intuitive balancing of the
impedance that addressed the effects of demand and service inflation
found in earlier FCA approaches (see Paez et al., 2019). This provides
more useful information because it does not assume that people are
limited to service within pre-defined boundaries (Paez et al., 2019).

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bicycle share equity has typically
used a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bicycle share demand (J.
Wang and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

\hypertarget{previous-research}{%
\subsection{2.4 Previous Research}\label{previous-research}}

Our analysis builds upon a previous study (Hosford and Winters, 2018),
which found that areas in Hamilton with less advantage are better served
by the city's PBSP compared to other Canadian cities (i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
deprived have greater access. Hosford and Winters (2018, pp.~)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton by default covers more of the
less advantaged areas in the city. However, there is also a great deal
of variation in income in the city center because of the local
university and increasing gentrification. Hosford and Winters (2018)
took a macro-level approach in their analysis by using dissemination
areas \emph{across} the city as the unit of analysis. They did not focus
specifically on the core service area and did not differentiate between
conventional and equity stations.

Therefore, this paper contributes to the cycling literature by expanding
the analysis of Hosford and Winters (2018) to assess equity in
accessibility to HBS and explore the contribution of the ``equity''
stations added to Hamilton's PBSP.

@@

The case study of this paper is the city of Hamilton, Ontario, Canada.
Before June 2020, the program was known as Social Bicycles or SoBi
Hamilton, but is now called Hamilton Bike Share (HBS). The core service
area spans 40 sq.km of the city, although it was planned to be 20 sq.km
(Hamilton, 2015a), and roughly 138,000 people are within 30 minutes
walking of a station {[}see Figure
\ref{fig:hamilton-and-sobi-service-area}{]}. This represents roughly one
fifth of the total population of the Hamilton Census Metropolitan Area
according to the 2016 Canadian Census. The City of Hamilton undertook a
large public engagement campaign to validate the locations of stations
that had been selected and to crowdsource potential locations for
additional locations (Hamilton, 2014). Most of the suggested locations
were in the east end of the core service area that lack stations or in
neighborhoods not serviced by the PBSP. The program was enthusiastically
welcomed in the city in 2015 - within three weeks of launching, 10,000
trips had been made (Hamilton, 2015b), however inadequate coverage given
the size of Hamilton Bike Share's service area was identified as a
problem early on, and transportation planners noted that the small size
(i.e., supply of bicycle racks) and low quantity of stations would lead
to challenges in balancing the system (Hamilton, 2015a).
